---
title: "lab5-Data Wrangling"
author: "Ran Zhu"
format: html
editor: visual
embed-resources: true
fig-width: 6
fig-height: 4
theme: flatly
---

## 0.1. Setup

```{r}
library(data.table)
library(dtplyr)
library(dplyr)
library(leaflet)
```

## 0.2. Load the MET data

```{r}
# Load the meteorological data
met <- fread(
  "https://raw.githubusercontent.com/USCbiostats/data-science-data/master/02_met/met_all.gz"
)

# Convert met to a regular data frame
met <- as.data.frame(met)

# Download the station data
stations <- read.csv(
  "https://noaa-isd-pds.s3.amazonaws.com/isd-history.csv"
)

# Convert USAF station ID to integer
stations$USAF <- suppressWarnings(
  as.integer(stations$USAF)
)

# Deal with blanks and invalid station IDs
stations$USAF[stations$USAF == 999999] <- NA
stations$CTRY[stations$CTRY == ""] <- NA
stations$STATE[stations$STATE == ""] <- NA

# Select the relevant columns and keep unique records
stations <- unique(
  stations[, c("USAF", "CTRY", "STATE")]
)

# Drop records with missing USAF IDs
stations <- stations[!is.na(stations$USAF), ]

# Remove duplicated USAF IDs
stations <- stations[!duplicated(stations$USAF), ]
```

## 0.3. Merge the meteorological data with the station data

```{r}
met <- merge(
  x = met,
  y = stations,
  by.x = "USAFID",
  by.y = "USAF",
  all.x = TRUE,
  all.y = FALSE
)

# Inspect the merged data
dim(met)
head(met[, c("USAFID", "WBAN", "CTRY", "STATE")])
```

## Q1. Representative station for the US

```{r}
# Keep observations from the continental United States
met_us <- met |>
  filter(
    CTRY == "US",
    !is.na(STATE),
    !(STATE %in% c("AK", "HI"))
  )

# Calculate the average measurements for each weather station
station_avg <- met_us |>
  summarise(
    temp_avg = mean(temp, na.rm = TRUE),
    wind_avg = mean(wind.sp, na.rm = TRUE),
    pressure_avg = mean(atm.press, na.rm = TRUE),
    .by = c(USAFID, STATE)
  )

# Calculate the median of the station-level averages
median_temp <- quantile(
  station_avg$temp_avg,
  probs = 0.5,
  na.rm = TRUE
)

median_wind <- quantile(
  station_avg$wind_avg,
  probs = 0.5,
  na.rm = TRUE
)

median_pressure <- quantile(
  station_avg$pressure_avg,
  probs = 0.5,
  na.rm = TRUE
)
```

```{r}
# Identify the station closest to each median
temp_index <- which.min(
  abs(station_avg$temp_avg - median_temp)
)

wind_index <- which.min(
  abs(station_avg$wind_avg - median_wind)
)

pressure_index <- which.min(
  abs(station_avg$pressure_avg - median_pressure)
)

# Create a table containing the three representative stations
representative_stations <- data.frame(
  variable = c(
    "Temperature",
    "Wind speed",
    "Atmospheric pressure"
  ),
  
  USAFID = c(
    station_avg$USAFID[temp_index],
    station_avg$USAFID[wind_index],
    station_avg$USAFID[pressure_index]
  ),
  
  STATE = c(
    station_avg$STATE[temp_index],
    station_avg$STATE[wind_index],
    station_avg$STATE[pressure_index]
  ),
  
  median_value = c(
    median_temp,
    median_wind,
    median_pressure
  ),
  
  station_value = c(
    station_avg$temp_avg[temp_index],
    station_avg$wind_avg[wind_index],
    station_avg$pressure_avg[pressure_index]
  )
)

representative_stations
```

```{r}
n_distinct(representative_stations$USAFID)
```

The median station for temperature is USAFID 720458 in KY. Its mean temperature is 23.681730°C, which is only 0.002329°C away from the median temperature of 23.684059°C. For wind speed, the median station is USAFID 720929 in WI. Its mean wind speed is the same as the median wind speed, 2.461838 m/s. For atmospheric pressure, the median station is USAFID 722238 in AL. Its mean atmospheric pressure is 1014.690611 hPa, which is about 0.000537 hPa away from the median of 1014.691148 hPa. These are three different stations, so they do not coincide. These three representative stations do not coincide because they have different USAFIDs.

## Q2. Representative station per state

```{r}
station_coordinates <- met_us |>
  summarise(
    lat = mean(lat, na.rm = TRUE),
    lon = mean(lon, na.rm = TRUE),
    .by = c(STATE, USAFID)
  ) |>
  filter(
    !is.na(lat),
    !is.na(lon)
  )

state_median_coordinates <- station_coordinates |>
  summarise(
    median_lat = quantile(
      lat,
      probs = 0.5,
      na.rm = TRUE
    ),
    median_lon = quantile(
      lon,
      probs = 0.5,
      na.rm = TRUE
    ),
    number_of_stations = n(),
    .by = STATE
  ) |>
  arrange(STATE)

state_median_coordinates
```

The results show the median latitude and longitude of the stations in each state.

## Q3. In the middle?

```{r}
# Calculate the mean latitude and longitude for each state
state_mean_coordinates <- station_coordinates |>
  summarise(
    mean_lat = mean(lat, na.rm = TRUE),
    mean_lon = mean(lon, na.rm = TRUE),
    .by = STATE
  )

# Add each state's mean coordinates to its stations
station_distances <- merge(
  x = station_coordinates,
  y = state_mean_coordinates,
  by = "STATE",
  all.x = TRUE,
  all.y = FALSE
)

# Calculate each station's distance from its state mean coordinates
station_distances$distance_to_mean <- sqrt(
  (station_distances$lat - station_distances$mean_lat)^2 +
    (station_distances$lon - station_distances$mean_lon)^2
)

# Sort stations by state and distance
station_distances <- station_distances[
  order(
    station_distances$STATE,
    station_distances$distance_to_mean
  ),
]

# Keep the closest station in each state
closest_stations <- station_distances[
  !duplicated(station_distances$STATE),
]

closest_stations
```

```{r}
median_points <- data.frame(
  STATE = state_median_coordinates$STATE,
  USAFID = NA_integer_,
  lat = state_median_coordinates$median_lat,
  lon = state_median_coordinates$median_lon,
  point_type = "Median coordinates"
)

closest_station_points <- data.frame(
  STATE = closest_stations$STATE,
  USAFID = closest_stations$USAFID,
  lat = closest_stations$lat,
  lon = closest_stations$lon,
  point_type = "Station closest to mean"
)

map_points <- rbind(
  median_points,
  closest_station_points
)
```

```{r}
table(map_points$point_type)
nrow(map_points)
```

```{r}
# Create a color palette for the two types of points
point_colors <- colorFactor(
  palette = c("blue", "red"),
  domain = map_points$point_type
)

# Create the interactive map
leaflet(map_points) |>
  addTiles() |>
  addCircleMarkers(
    lng = ~lon,
    lat = ~lat,
    radius = 5,
    color = ~point_colors(point_type),
    fillColor = ~point_colors(point_type),
    fillOpacity = 0.8,
    label = ~paste0(
      STATE,
      ": ",
      point_type,
      ifelse(
        is.na(USAFID),
        "",
        paste0(", USAFID ", USAFID)
      )
    )
  ) |>
  addLegend(
    position = "bottomright",
    pal = point_colors,
    values = ~point_type,
    title = "Point type",
    opacity = 1
  )
```

The map contains 96 points, with two points for each of the 48 states. The blue points represent the median latitude and longitude calculated in Question 2, while the red points represent the actual stations closest to the mean latitude and longitude. Some red and blue points are very close, while others are farther apart because the mean and median coordinates are not always the same.

## Q4. Means of means

```{r}
# Calculate the average temperature for each state
state_temperature <- met_us |>
  summarise(
    state_mean_temp = mean(temp, na.rm = TRUE),
    .by = STATE
  ) |>
  arrange(STATE)
```

```{r}
# Classify each state based on its average temperature
state_temperature$temp_category <- ifelse(
  state_temperature$state_mean_temp < 20,
  "low",
  ifelse(
    state_temperature$state_mean_temp < 25,
    "Mid",
    "High"
  )
)

# Show the states and their temperature categories
state_temperature

# Show the number of states in each category
table(state_temperature$temp_category)
```

```{r}
# Merge the state categories back into the observation-level data
met_classified <- merge(
  x = met_us,
  y = state_temperature,
  by = "STATE",
  all.x = TRUE,
  all.y = FALSE
)

# Generate the summary table for the three temperature categories
q4_summary <- met_classified |>
  summarise(
    number_of_states = n_distinct(STATE),
    number_of_entries = n(),
    number_of_NA_entries = sum(is.na(temp)),
    number_of_stations = n_distinct(USAFID),
    mean_temperature = mean(temp, na.rm = TRUE),
    mean_wind_speed = mean(wind.sp, na.rm = TRUE),
    mean_atmospheric_pressure = mean(
      atm.press,
      na.rm = TRUE
    ),
    .by = temp_category
  )

# Arrange the categories in the order shown in the question
q4_summary$temp_category <- factor(
  q4_summary$temp_category,
  levels = c("low", "Mid", "High")
)

q4_summary <- q4_summary |>
  arrange(temp_category)

# Round the mean values to two decimal places
q4_summary <- q4_summary |>
  mutate(
    mean_temperature = round(
      mean_temperature,
      2
    ),
    mean_wind_speed = round(
      mean_wind_speed,
      2
    ),
    mean_atmospheric_pressure = round(
      mean_atmospheric_pressure,
      2
    )
  )

# Display the final summary table
q4_summary
```

There are 11 states in the low-temperature group, 25 states in the Mid-temperature group, and 12 states in the High-temperature group. The Mid group has the largest number of records, missing temperature values, and weather stations. The mean temperatures are 19.0°C, 22.4°C, and 27.8°C for the low, Mid, and High groups, respectively. The mean wind speeds and atmospheric pressures are relatively similar across the three groups.
